UMAP reveals cryptic population structure and phenotype heterogeneity in large genomic cohorts

### 2018-12-11 ms/main.tex
small edits to summary

%Authors may benefit from consulting with a science writer or press officer to ensure they effectively communicate their findings to a general audience.

Population structure is a systematic difference in allele frequencies between populations. It is a consequence of non-random mating from factors such as geographic isolation and founder effects from migrations, and is a common confounder in genetic studies. We use a newly developed method (UMAP) to visualize population structure from three genomic datasets. Using genotype data alone, we reveal numerous patterns that relate population structure and ancestry to traits such as white blood cell count, height, and FEV1, a baseline measure for airway obstruction. We demonstrate that UMAP reveals previously unobserved patterns and fine-scale structure, and that when working in higher dimensions it can be used to study, for example, the spatial distribution of genetic relatedness\adpcomment{ (talking about the UK map here)}. We show that visualizations work especially well in large datasets containing populations with diverse backgrounds, which are rapidly becoming more common, and that unlike other visualization methods, we can preserve intuitive connections between populations that reflect their shared ancestries. The combination of these results and the effectiveness of the strategy on large and diverse datasets make this an important approach for exploratory analysis for geneticists studying ancestral events and phenotype distributions, and the images generated will pique the interest of general audiences.

>>>>

%Authors may benefit from consulting with a science writer or press officer to ensure they effectively communicate their findings to a general audience.
Because of geographic isolation, individuals tend to be more genetically related to people living nearby than to people living far. This is an example of population structure, a situation where a large population contains subgroups that share more than the average amount of DNA. This structure can tell us about human history, and it can also have a large effect on medical studies.  We use a newly developed method (UMAP) to visualize population structure from three genomic datasets. Using genotype data alone, we reveal numerous subgroups related to ancestry and correlated with traits such as white blood cell count, height, and FEV1, a baseline measure for airway obstruction. We demonstrate that UMAP reveals previously unobserved patterns and fine-scale structure. We show that visualizations work especially well in large datasets containing populations with diverse backgrounds, which are rapidly becoming more common, and that unlike other visualization methods, we can preserve intuitive connections between populations that reflect their shared ancestries. The combination of these results and the effectiveness of the strategy on large and diverse datasets make this an important approach for exploratory analysis for geneticists studying ancestral events and phenotype distributions, and the images generated will pique the interest of general audiences.
